---
title: "Introduction to terraces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to terraces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%",
  warning = FALSE
)
```

The `terraces` package provides aggregation-consistent disaggregation methods for `terra` rasters. Unlike standard bilinear interpolation, when applied to a coarse raster, these methods produce a fine-scale result that can be re-aggregated to recover the original coarse values. See the [README](https://matthewkling.github.io/terraces/) for background, and the [method-comparison article](https://matthewkling.github.io/terraces/articles/method-comparison.html) for a more detailed exploration of how these methods work on real elevation data.

This vignette walks through basic usage.

```{r setup}
library(terra)
library(terraces)
```


## Disaggregating a coarse raster

We'll use the `volcano` dataset that ships with base R as a coarse input:

```{r data, fig.height = 3.5}
coarse <- rast(volcano) |> aggregate(5) |> trim()
plot(coarse, main = "Coarse input")
```

Each of the three methods takes the coarse raster and a disaggregation factor:

```{r disagg, fig.height = 3.5}
fine_bl  <- disagg_bl( coarse, fact = 5) |> setNames("disagg_bl")
fine_cub <- disagg_cub(coarse, fact = 5) |> setNames("disagg_cub")
fine_pyc <- disagg_pyc(coarse, fact = 5) |> setNames("disagg_pyc")

plot(c(fine_bl, fine_cub, fine_pyc), nr = 1)
```

The three methods produce qualitatively similar output but differ in their smoothness priors. See the README or the method-comparison article for guidance on choosing between them.


## Mapping the boundary band

The pre-sharpening methods (`disagg_bl()` and `disagg_cub()`) have a small boundary band where mass preservation degrades slightly. You can map this affected zone with `edge_effects()`:

```{r edge, fig.height = 3.5}
ee <- edge_effects(coarse, fact = 5, method = "cubic")
plot(ee, main = "Edge-effect zone (cubic method)")
```

Values 1 and 2 indicate cells affected by the pre-sharpening filter and the boundary interpolation fallback, respectively. For studies that care about precision near a particular region of interest, the cleanest practical workaround is to disaggregate over a larger domain that extends beyond the study area, then crop the result — this places the edge effects safely outside the cells you actually use. Alternatively, if you need exact mass preservation everywhere — for example, when tiling mosaics or when extending the domain isn't possible — use `disagg_pyc()`, which can be slower for large rasters but isn't subject to these effects.


## Verifying mass preservation

Re-aggregating the fine output should recover the original coarse values. 

To compare methods fairly, we'll exclude the boundary band where the pre-sharpening methods are known to have degraded accuracy. We build a coarse-resolution mask flagging any coarse cell that contains a level-2 boundary fine cell:

```{r mask}
coarse_mask <- aggregate(ee, 5, fun = "max")

max_error <- function(fine, coarse, mask){
      back <- aggregate(fine, fact = 5, fun = "mean")
      max(abs(values(mask(coarse - back, mask, maskvalues = 2))), na.rm = TRUE)
}
```

For `disagg_pyc()`, residuals are at machine precision — mass is preserved exactly by construction, with or without the mask:

```{r verify-pyc}
max_error(fine_pyc, coarse, coarse_mask)
```

For `disagg_bl()` and `disagg_cub()`, interior residuals are very small (~6 cm on a DEM covering a 100 m range), but are nonzero due to finite-radius kernel truncation:

```{r verify-bl}
max_error(fine_bl, coarse, coarse_mask)
```

For comparison, here's the same check with standard bilinear disaggregation using `terra::disagg()`. Residuals here are orders of magnitude larger — standard bilinear isn't mass-preserving:

```{r verify-std}
fine_std <- terra::disagg(coarse, fact = 5, method = "bilinear")
max_error(fine_std, coarse, coarse_mask)
```



## Learn more

- The [README](https://matthewkling.github.io/terraces/) explains the package's design and helps you choose between methods.
- The [method-comparison article](https://matthewkling.github.io/terraces/articles/method-comparison.html) works through a detailed analysis with a real DEM, comparing methods based on their assumptions, accuracy, and computational speed.
- Function help: `?disagg_bl`, `?disagg_cub`, `?disagg_pyc`, `?edge_effects`.
